Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-13 kappa.tex
Rework results and conclusions with Em Yeu.

This doctoral dissertation consists of a cover summary and the following
papers, which are referred to by their Roman numerals in the text.

@@

% ----------------------------------------------------------------------
\subsection{Constant-climate runs}
% ----------------------------------------------------------------------

\begin{figure*}
  \centering
  \vspace{-16mm}
  \makebox[0pt]{
  \begin{minipage}[t]{112.5mm}\vspace{0pt}
    \includegraphics{plot-lgm}
  \end{minipage}
  \hspace{1cm}
  \begin{minipage}[t]{47.5mm}\vspace{0pt}
    \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
             an horizontal resolution of 3\,km. Modelled basal topography,
             cold-based areas (hatches), surface topography (100\,m contours)
             and surface velocities. The simulation is forced by atmospheric
             fields from the NARR \citep{Mesinger.etal.2006} and a temperature
             depression of 7\,\degC, using the set-up described in \CCLI. There
             are several issues with this run. First, all east-facing margins
             are too extensive as compared to available reconstructions.
             Second, marine advance on the Pacific Shelf is too restrictive,
             most notably in the south-western corner where the Puget and Juan
             de Fuca lobes are missing. Third, en-glacial temperatures are
             anomalously low due to brutal application of cold climate
             condition to ice-free bedrock at the inception of the run.
             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
  \end{minipage}
  }
  \vspace{-16mm}
  \label{fig:plot-lgm}
\end{figure*}

@@

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdeffect}}
  \caption{Effect of temperature variability on Cordilleran ice sheet
           simulations. Modelled basal topography and surface topography
           (250\,m contours). The simulations use atmospheric forcing from the
           NARR \citep{Mesinger.etal.2006}, a temperature depression of
           7\,\degC, and different treatments of temperature variability
           including
           \textbf{(a)} no daily variability, $\sigma=0$;
           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged 
           standard deviation value, $\sigma=3.09$\,K;
           \textbf{(c)} monthly standard deviation grids derived from the NARR
           and
           \textbf{(d)} monthly grids for ice-free terrain and the
           Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
           parametrisation in \PSDP, ${\sigma=-0.15\cdot{T_m}+1.66}$.
           Differences in surface elevation are displayed relative to case (c).
           Note the different colour scales.}
  \label{fig:plot-sdeffect}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Transient-climate runs}
% ----------------------------------------------------------------------

@@

% ======================================================================
\section{Future perpectives}
% ======================================================================
% Model improvements
% Fieldwork
% Geomorphological-oriented plots

%\begin{figure}
%  \centering
%  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
%  \caption{The Tuya Lake Moraine in northern British Columbia.}
%  \label{fig:photo-tuya-moraine}
%\end{figure}

% References
\newcommand{\urlprefix}[0]{}  % remove default "URL" prefix
\bibliographystyle{copernicus}
\bibliography{refs/references.bib}

>>>>

This doctoral dissertation consists of a cover summary and the following
journal articles, which are referred to by their Roman numerals in the text.

@@

% ----------------------------------------------------------------------
\subsection{Constant-climate simulations}
% ----------------------------------------------------------------------

\noindent\textbf{\CCLI} --- \bibentry{Seguinot.etal.2014}
\bigskip

As a first step towards simulations of the Cordilleran ice sheet through the
last glacial cycle, I have chosen to run the ice sheet model under a constant
climate, targeting the LGM as the period for which ice sheet limits are best
known. These simulations, presented in \CCLI, use constant temperature offsets
from the present climate state. They start from ice-free conditions and run for
10\,ka, a period chosen as representative of the ice-sheet build-up from
assumed, negligible ice cover. Although this assumption is disproved in \CCYC,
the simplistic set-up used in \CCLI\ allowed to test model sensitivity to the
choice of present-day climate forcing data, without involving too many
additional parameters.

Discrepancies between input data used (\cref{tab:datasources}) caused much
variability in modelled ice sheet geometries (\CCLI, Figs.~5--7). In
particular, continental parts of the ice-sheet are strongly dependent on input
precipitation patterns, which are often poorly resolved by atmospheric
reanalysis fields used to force the model (\CCLI, Figs.~10--14). The closest
match between model output and the reconstructed ice margin \citep{Dyke.2004}
was obtained using atmospheric forcing from the North American Regional
Reanalysis \citep[NARR,][]{Mesinger.etal.2006} and a temperature depression of
7\,\degC\ (\CCLI, Fig.~15). The relevant simulation is reproduced here
(\cref{fig:plot-lgm}), using a higher horizontal resolution of 3\,km, instead
of 10\,km as in the paper. Following this sensitivity study, I have retained
the NARR to force subsequent simulations of the Cordilleran ice sheet
presented here and in \CCYC.

\begin{figure*}
  \centering
  \vspace{-16mm}
  \makebox[0pt]{
  \begin{minipage}[t]{112.5mm}\vspace{0pt}
    \includegraphics{plot-lgm}
  \end{minipage}
  \hspace{1cm}
  \begin{minipage}[t]{47.5mm}\vspace{0pt}
    \caption{Constant-climate simulation of the LGM Cordilleran ice sheet using
             an horizontal resolution of 3\,km. Modelled basal topography,
             cold-based areas (hatches), surface topography (100\,m contours)
             and surface velocities. The simulation is forced by atmospheric
             fields from the NARR \citep{Mesinger.etal.2006} and a temperature
             depression of 7\,\degC, using the set-up described in \CCLI. There
             are several issues with this run. First, all east-facing margins
             are too extensive as compared to available reconstructions.
             Second, marine advance on the Pacific Shelf is too restrictive,
             most notably in the south-western corner where the Puget and Juan
             de Fuca lobes are missing. Third, en-glacial temperatures are
             anomalously low due to brutal application of cold climate
             condition to ice-free bedrock at the inception of the run.
             \todo{Add black frame, and maybe the LGM outline from Dyke?}}
    \label{fig:plot-lgm}
  \end{minipage}
  }
  \vspace{-16mm}
\end{figure*}

@@

\noindent\textbf{\PSDV} --- \bibentry{Seguinot.2013}
\bigskip

\noindent\textbf{\PSDP} --- \bibentry{Seguinot.Rogozhina.2014}
\bigskip

Next to the atmospheric forcing, another critical source of uncertainty in
palaeo-ice sheet modelling is the choice of parameters used by the surface mass
balance model \citep[e.g.,][]{Hebeler.etal.2008}. In our case, these parameters
are the degree-day factors for snow, $F_s$, and ice, $F_i$, and the standard
deviation of near-surface air temperature, $\sPDD$. Although the first two can
only be deduced from laborious surface mass-balance measurements on present-day
glaciers \citep[e.g.,][]{Shea.etal.2009}, $\sPDD$ can be estimated from
temperature time-series \citep[e.g.,][]{Fausto.etal.2011}.

In \PSDV, I used near-surface air temperatures from the ERA-Interim reanalysis
\citep{Dee.etal.2011} to compute a global, monthly distribution of $\sPDD$.
The resulting grids show
that, although often assumed constant, $\sPDD$ varies greatly by location and
through the year, most particularly in the polar regions
(\cref{fig:plot-sdmap}; \PSDV, Fig.~1). Using a simple surface mass balance
model, I then demonstrated the effect of these variations on the computation of
positive degree-days (\PSDV, Fig.~2), and surface mass-balance (\PSDV, Fig.~3).

The analysis is taken further in \PSDP, using re-analysed air temperatures from
ERA-40 \citep{Uppala.etal.2005}. Although of coarser spatial resolution, the
ERA-40 time series have a longer time span than the ERA-Interim ones, which
is necessary to compute standard deviation values that are representative of
the long-term daily variability \citep{Rogozhina.Rau.2014}. In \PSDP, we
demonstrated that, over the Greenland ice sheet, there exist an inverse
correlation between $\sPDD$ and monthly mean temperatures, $T_m$
(\PSDP, Fig.~1). This correlation could potentially be used as a
parametrisation, and can be expressed as (\PSDP, Eq.~5)

\begin{equation}
    \label{eqn:sigma}
    \sPDD = -0.15 \cdot T_{m} + 1.66\,.
\end{equation}

The effect of the PDD standard deviation, $\sPDD$, on simulations of the
Cordilleran ice sheet is not shown in any of the papers. However, following
these two studies, a natural step forward was to implement, in PISM, spatially
and seasonally variable $\sPDD$ input (PISM issue \issue{179}), and the
parametrisation expressed by \cref{eqn:sigma} over ice-filled grid cells
(\issue{265}). Their effect on the modelled Cordilleran ice sheet surface is
demonstrated here (\cref{fig:plot-sdeffect}) using a constant-climate set-up
similar to the one used in \CCLI.

@@

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{plot-sdeffect}}
  \caption{Effect of temperature variability on Cordilleran ice sheet
           simulations. Modelled basal topography and surface topography
           (250\,m contours). The simulations use atmospheric forcing from the
           NARR \citep{Mesinger.etal.2006}, a temperature depression of
           7\,\degC, and different treatments of temperature variability
           including
           \textbf{(a)} no daily variability, $\sPDD=0$;
           \textbf{(b)} a constant, mean summer (JJA), model domain-averaged
           standard deviation value, $\sPDD=3.09$\,K;
           \textbf{(c)} monthly standard deviation grids derived from the NARR
           and
           \textbf{(d)} monthly grids for ice-free terrain and the
           Greenland ice sheet, ERA-40-based \citep{Uppala.etal.2005}
           parametrisation in \PSDP, ${\sPDD=-0.15\cdot{T_m}+1.66}$.
           Differences in surface elevation are displayed relative to case (c).
           Note the different colour scales.}
  \label{fig:plot-sdeffect}
\end{figure}

% ----------------------------------------------------------------------
\subsection{Glacial cycle simulations}
% ----------------------------------------------------------------------

\noindent\textbf{\CCYC} --- \bibentry{Seguinot.etal.inpreparation}
\bigskip

From the beginning, the primary goal of this thesis has been to perform
simulations of the Cordilleran ice sheet through the last glacial cycle. This
is the topic of \CCYC, where monthly air temperature, precipitation, and PDD
standard deviation climatologies from the NARR are amended by
palaeo-temperature time-series from six proxy records to produce a
time-dependent climate forcing (\CCYC, Fig.~2). The resulting simulations all
produce two major glaciations between MISs~4 and 2. However the differences in
timing and magnitude of these events is considerable (\CCYC, Fig.~3--4).

Geological evidence regarding the timing of glaciation allowed us to select two
simulations, driven by the GRIP \citep{Dansgaard.etal.1993} and EPICA
\citep{Jouzel.etal.2007} ice-core records, as producing more realistic
reconstructions of the last glacial cycle. I have re-run these two simulations
at a higher resolution of 6\,km, to allow for a more detailed study of
transient ice cap cover before and after the LGM, stability of major
ice-dispersal centres, and patterns of deglaciation (\CCYC, Fig.~5--15).
Modelled ice volume evolution and snapshots of ice flow patterns from one of
these two simulations are reproduced here (\cref{fig:plot-snapshots}).

\todo{Verify that figure numbers remain correct as \CCYC\ evolves.}

@@

% ======================================================================
\section{Conclusion and perpectives}
% ======================================================================

In this thesis, I present numerical reconstructions of the former Cordilleran
ice sheet to a level of detail that has, to my knowledge, never been attained
before. The coupled SIA-SSA approximation is crude, yet it is today the only
approach that allows, combined with massively parallel computing, high-spatial
resolution over the multi-millenial time-scales of ice sheet growth and decay.

Through sensitivity studies, I have identified the predominant sources of
uncertainty, which I have assessed using a combination of detailed
meteorological datasets and available geological evidence. The simulations
supporting this thesis allowed estimates of the LGM Cordilleran ice sheet
extent and volume, identification of the main ice-dispersal centres, and
partial understanding of the deglaciation patterns. However, there remains much
room for model improvements and further analysis of the simulation output.

% ----------------------------------------------------------------------
\subsection{Precipitation change}
% ----------------------------------------------------------------------

Throughout the thesis, precipitation rates are prescribed as monthly, spatial
grids, but assumed equal to their present values. Although the distribution of
precipitation has been undoubtedly affected by past changes in global climate
and the intermittent presence of ice sheets and ice caps in the Northern
Hemisphere, its changes are difficult to assess. In the North American
Cordillera, much of the continental interior was likely dryer than today due to
the presence of ice caps reinforcing the topographic barrier formed by the
Pacific Coast ranges. However, some regions might have become rainier as a side
effect of redirected winds.

In meteorology, these effects are typically analysed using atmospheric
circulation models. However, I have shown in \CCLI\ that atmospheric reanalyses
which are circulation models that use data assimilation techniques, fail to
agree on the present-day pattern of precipitation in the Cordillera. The
discrepancies, due to highly mountainous terrain, have dramatic effects on the
ice-sheet-model response. Considering additional difficulties inherent to
modelling past climates, and the lack of data for assimilation, we can only
expect that palaeo-climate reconstructions will produce even higher
discrepancies in the ice-sheet-model response.

% ----------------------------------------------------------------------
\subsection{Geothermal heat flux}
% ----------------------------------------------------------------------

Geothermal heat flux, controlling en-glacial temperature distribution and the
location of basal melt areas, was assumed constant. However, the North America
is an active subduction zone with complex geothermal patterns
\citep{Blackwell.Richards.2004}. Including this variability in the ice-sheet
model will offer new opportunities to study its output.

Most notably, modelled basal flow could be compared with glacial lineations,
one of the main sources of geomorphological evidence used to reconstruct former
ice sheets. For instance, in our simulations, large portions of the Cordilleran
ice sheet remain cold-based throughout deglaciation (\CCYC, Fig.~15). As
cold-based glaciers typically do not deform their beds, this could indicate
that some of the lineation patterns we observe in the North American Cordillera
could pre-date deglaciation. Instead, they may have been formed during an
earlier phase of advance or retreat of the ice sheet.

% ----------------------------------------------------------------------
\subsection{The Cassiar ice cap}
% ----------------------------------------------------------------------

In the summer of 2012, I have co-organised and participated in a fieldwork with
Martin Margold and Ping Fu to collect samples of glacial deposits in the
southern Cassiar Mountains of northern British Columbia
(\cref{fig:map-cordillera}). The primary target of the fieldwork was a
triple-crested, sharp, ca. 50\,m-high moraine located in the vicinity of Tuya
Lake on the western slope of the Cassiar Mountains
(\cref{fig:photo-tuya-moraine}). Although this is a unique feature in the area,
the size of the moraine and its location at the foothill of the mountain range
indicate that it was likely formed by a tributary glacier from an ice cap,
rather than a stand-alone valley glacier. The relatively low-elevated Cassiar
Mountains do not appear as a major nucleation centre in our simulations (\CCYC,
Fig.~5). Therefore, it is tempting to think that such a large feature may
result from the late-glacial re-advance of a remnant of the decaying ice sheet.
High-resolution modelling of the deglaciation phase, perhaps at a regional
scale, may help to study this hypothesis.

\begin{figure}
  \centering
  \makebox[0pt]{\includegraphics{photo-tuya-moraine}}
  \caption{The Tuya Lake Moraine in northern British Columbia.}
  \label{fig:photo-tuya-moraine}
\end{figure}

% References
\renewcommand{\urlprefix}[0]{}  % remove default "URL" prefix
\bibliographystyle{copernicus}
\bibliography{refs/references.bib}
